---
title: "602Project_part1"
author: "Ze Yu"
date: "2024-02-09"
output: word_document
---
```{r analysis, include=FALSE}
library(ggplot2)
library(dplyr)
library(mosaic)
library(resampledata)
```
Part1:
***Read data***:
```{r echo=TRUE}
# Read data 
data = read.csv("bondsdata.csv")
# Remove year 2001
data <- data[-c(15), ]
data
```

***Build Model***:
```{r echo=TRUE}
ggplot(data, aes(x = season, y = hrat)) + geom_smooth(method = "lm") + geom_point() + xlab("Season") + ggtitle("home run rate vs season") 
predicted_rate = lm(hrat ~ season, data=data)
predicted_rate$coef
```
The linear regression equation representing the model is $\hat{HRAT_{i}} = -7.992499290 + 0.004044169Year_{i}$.
***Correlation Coefficient Check***:
```{r echo=TRUE}
season <- data$season
hrat <- data$hrat
cor(season, hrat)
```
Correlation Coefficient shows that there is a strong positive correlation between the season/year and home run rate of Barry Bonds.

***Check significance of coefficient estimates***
Null hypothesis: $H_{0}: \beta_{1} = 0$, $H_{0}: \beta_{0} = 0$

Alternative hypothesis: $H_{A}: \beta_{1} \neq 0$, $H_{0}: \beta_{0} \neq 0$

We will set the alpha value to 0.05.

We can use a t test to check our claim.
```{r echo=TRUE}
summary(predicted_rate)
```

From our t-test, we get test statistics of $\beta_{0}$ is -4.549782 and $\beta_{1}$ is 4.589384. P-values of $\beta_{0}$ is 0.0006664296 and $\beta_{1}$ is 0.0006222474. The p-values are smaller than the set alpha value of 0.05, so we reject our null hypothesis that the linear regression coefficient and intercept are 0. Therefore, we can conclude that the home run rate of Bary Bonds can be expressed as a linear function of the season, and since $\beta_{1} > 0$, we can say it is also positive. 

We also get an R-squared value of 0.6371, meaning out independant variable explains approx. 63.71% of the variance in the dependent variable, which is quite good.


***Residual Analysis***:

There are two conditions that must be met for our linear regression model to be valid. 

**1. Normality of residuals:** The dependent variable (hrat) must be normally distributed with a mean of $\mu$ and standard deviations of $\sigma$. To check this we will plot a stat_qq plot of the residuals since $e_{i} = y_{i} - \hat{y_{i}}$, if y is normally distributed, so will the residuals.

**2. Homoscedasticity:** For each distinct value of the independant variable (season), the dependent variable (hrat) has the same standard deviation $\sigma$. To check this, we will plot a scatterplot of the fitted values and the residuals.
```{r echo=TRUE}
# Get the and residuals fitted values
predicted.rate = predicted_rate$fitted.values
ei_hrat = predicted_rate$residuals  
data.df = data.frame(predicted.rate, ei_hrat)
```


**Normality of Residuals Plot:**
```{r echo=TRUE}
ggplot(data.df, aes(sample = ei_hrat)) +  stat_qq(col='blue') + stat_qqline(col='red') + ggtitle("Normal Probability Plot of Residuals")
```
Looking the normality probability plot, the residuals do seem to be approximately normally distributed and therefore so is the  dependent variable (hrat). The normality of residuals condition holds.

**Homoscedasticity:**
```{r echo=TRUE}
ggplot(data.df, aes(x = predicted.rate, y = ei_hrat)) +  geom_point(size=2, col='blue', position="jitter") + xlab("Predicted Home Run Rate") + ylab("Residuals") + ggtitle("Plot of Fits to Residuals") + geom_hline(yintercept=0, color="red", linetype="dashed")
```
Looking the plot of fits to residuals, the residuals do seem to be evenly distributed over the home run rate. We can they say that the condition of homoscedasticity holds.

Since both conditions hold, our linear regression model is valid.

***Predict home run rate in season 2001***:
```{r echo=TRUE}
predict(predicted_rate, newdata=data.frame(season = 2001), interval="predict")
```

Using our linear regression model, we get a predicted HRAT of 0.099883 for Barry Bonds' 2001 season. Barry Bonds' actual HRAT for the 2001 season was 0.153400, so our predicted error is $0.153400-0.099883=0.053517$.

We also calculated a 95% prediction interval of HRAT for the 2001 season. This means we can say with 95% confidence that the value of Barry Bonds' HRAT for the 2001 season is between 0.06663 and 0.13314 based on our previous records. However, Barry Bonds' actual HRAT is not in this range. 

In conclusion, we cannot say whether or not Barry Bonds was on steroids in the 2001 season, but it does seem like there was a factor that attributed to a much larger HRAT in the 2001 season than what was expected.
